Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model

### 2022-07-28 Rmarkdown/paper.Rmd
added responses to reviewers

<!-- to do: check figure order -->
<!-- remove figure 9 from figures.R-->

@@

1. An average representing each pipe.
2. A smooth seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (this component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it).
3. A smooth global trend describing multi-year variation in the parameter of interest across all pipes.
4. A set of smooth pipe-specific trends that describe the deviations from the global multi-year trend particular to each pipe.
5. An autocorrelated error term.

Separation of the time-trend into multi-year and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the multi-year trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are long data gaps in the series representing one or more pipes. In these instances, the model "borrows" information from the global term. The autocorrelated error model improves the accuracy of parameter distributions and predictions, while helping to prevent overfitting [@simpson_modelling_2018]. And while time is the only covariate in our model, additional predictors can easily be added as smooth or linear functions.

@@

Lead, iron, aluminum, and manganese were measured by ICP-MS [@sm_3125] with method detection limits estimated at 0.2, 5, 5, and 2 µg L^-1^, respectively. An aliquot of each sample was filtered at 0.45 µm using a polytetrafluoroethylene (PTFE) filter and a vacuum filter apparatus. Losses of dissolved lead by adsorption to PTFE filters have been documented previously [@minning_systematic_2015], which here would tend to bias dissolved concentrations low and particulate concentrations high. In hard waters such as the feedwater for the model system, this bias is likely to be much lower than in soft waters.

@@

$$(2)~~f(x_t) = \sum_{j=1}^kb_j(x_t)\beta_j$$
$b_j(x_t)$ are the functions of $x_t$ on which $y$ is regressed, $\beta_j$ are the model coefficients, and $\epsilon_t$ is the error term. We fitted the example GAM in Figures 2--3 using the R package *mgcv*, which uses penalized likelihood maximization instead of ordinary least squares to choose coefficients while protecting against overfitting.

@@

$$(3)~~\epsilon_t = \phi\epsilon_{t-1} + \omega_t$$
where $\epsilon_t$ is the residual at time $t$, $\phi$ is the autoregressive parameter, and $\omega_t$ is an independent draw from a probability distribution.

First-order autocorrelation can be approximated as the correlation between the lagged and unlagged residuals, $\epsilon_t$ and $\epsilon_{t-1}$. A reasonable estimate of $\phi$ for the linear model residuals in Figure 2b, then, would be *r*. Unmodeled positive autocorrelation, as in Figure 2b-c, overstates the precision of parameter estimates, and so it should be either eliminated via model selection or taken into account explicitly. We chose to analyze the full length of each time series---including the conditioning period---to best estimate the autocorrelation structure.

@@

The models included one-dimensional global smoothers describing multi-year variation in lead across all time series (Figure 4b). The model of dissolved lead also included a global smoother describing seasonal variation (Figure 4a). Both models included group-level smoothers describing multi-year variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed elsewhere [@pedersen_hierarchical_2019]. The multi-year smoothers were thin-plate regression splines and the seasonal smoother was a cyclic cubic regression spline [@mgcv]. To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance. While omitted here, total lead could be predicted using the same model structure.

@@

where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$.

@@

$$
(7)~~
\begin{align}
generalized~additive~model~(GAM): \\
f_j(t) = X_j\beta_j + Z_jb_j \\
priors~for~GAM~coefficients: \\
\beta_j \sim T(0, 2.5, 3)~~[unpenalized] \\
b_j \sim N(0, \sigma_b)~~[penalized] \\
\sigma_b \sim N(0, 0.5) (seasonal~smoother) \\
\sigma_b \sim T(0, 2.5, 3) (global,~group-level~smoothers) \\ 
\end{align}
$$

where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function associated with the global multi-year trend). The priors on $\sigma_b$ help to prevent overfitting by penalizing complexity in the fitted smooth terms [@mgcv]. And while the chosen priors did help with model convergence, they did not exert a strong influence on the posterior (Figure S2); the likelihood---representing `r nrow(model_in)` observations in each model---was overwhelmingly influential. A summary of the fitted models, including convergence statistics and effective sample sizes, is provided in Figure S3 and S4.

@@

## Modeling autocorrelation and assessing performance

@@

Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant. This general pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation. The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to nonseasonal changes in water quality over the study.

@@

The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the large difference in lead release between the 0 and 1 mg P L^-1^ dose. The effect of orthophosphate dose was complex: lead release at 1 mg P L^-1^ was lower than at 0.5 mg P L^-1^ but higher than at 0.75 mg P L^-1^. We attribute the superior performance of the 0.75 mg P L^-1^ dose to an initial high dose of 2 mg P L^-1^. On the other hand, the initial 2 mg P L^-1^ dose caused excess lead release, highlighting an important trade-off that should be considered in applying a passivation-maintenance (high-low) dosing strategy. And since high orthophosphate doses may generate small lead-orthophosphate colloids/nanoparticles that are difficult to remove from drinking water [@lytle_lead_2020], the 1 mg P L^-1^ dose may be a better option in this case. The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring. This seasonal pattern may have multiple drivers: iron and water temperature, for instance, were also seasonal.

>>>>

<!-- to do: check figure order -->

@@

1. An average representing each pipe (all other components are centered on zero).
2. A smooth seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (this component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it).
3. A smooth global trend describing multi-year variation in the parameter of interest across all pipes.
4. A set of smooth pipe-specific trends that describe the deviations from the global multi-year trend particular to each pipe.
5. An autocorrelated error term.

Separation of the time-trend into multi-year and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the multi-year trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are long data gaps in the series representing one or more pipes. In these instances, the model "borrows" information from the other time series. The autocorrelated error model improves the accuracy of parameter distributions and predictions, while helping to prevent overfitting [@simpson_modelling_2018]. And while time is the only covariate in our model, additional predictors can easily be added as smooth or linear functions.

@@

Lead, iron, aluminum, and manganese were measured by ICP-MS [@sm_3125] with method detection limits estimated at 0.2, 5, 5, and 2 µg L^-1^, respectively. An aliquot of each sample was filtered at 0.45 µm using a polytetrafluoroethylene (PTFE) filter and a vacuum filter apparatus. Losses of dissolved lead by adsorption to PTFE filters have been documented previously [@minning_systematic_2015], which here would tend to bias dissolved concentrations low and particulate concentrations high. In hard waters such as the feedwater for the model system, this bias is likely to be much lower than in soft waters [@minning_systematic_2015].

@@

$$(2)~~f(x_t) = \sum_{j=1}^kb_j(x_t)\beta_j$$
$b_j(x_t)$ are the functions of $x_t$ on which $y$ is regressed, $\beta_j$ are the model coefficients, and $\epsilon_t$ is the error term. We fitted the example GAM in Figures 2--3 using the R package *mgcv* [@mgcv], which uses penalized likelihood maximization instead of ordinary least squares to choose coefficients while protecting against overfitting.

@@

$$(3)~~\epsilon_t = \phi\epsilon_{t-1} + \omega_t$$
where $\epsilon_t$ is the residual at time $t$, $\phi$ is the autoregressive parameter, and $\omega_t$ is an independent draw from a continuous---typically Gaussian---probability distribution.

First-order autocorrelation can be approximated as the correlation between the lagged and unlagged residuals, $\epsilon_t$ and $\epsilon_{t-1}$. A reasonable estimate of $\phi$ for the linear model residuals in Figure 2b, then, would be *r*. Unmodeled positive autocorrelation, as in Figure 2b-c, overstates the precision of parameter estimates, and so it should be either eliminated via model selection or taken into account explicitly. Here, we chose to analyze the full length of each time series---including the conditioning period---to best estimate the autocorrelation structure.

@@

The models included one-dimensional global smoothers describing multi-year variation in lead across all time series (Figure 4b). The model of dissolved lead also included a global smoother describing seasonal variation (Figure 4a). And while the model assumes that seasonality strength does not vary over time, some previous work has suggested that orthophosphate may damp seasonal variation in lead [@colling_measurement_1987]. This effect could be incorporated by allowing the seasonal smoother to vary with orthophosphate dose.

Both models included group-level smoothers describing multi-year variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed elsewhere [@pedersen_hierarchical_2019]. The multi-year smoothers were thin-plate regression splines and the seasonal smoother was a cyclic cubic regression spline [@mgcv]. To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance. And while it is omitted here, total lead could be predicted using the same model structure as dissolved lead.

@@

where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$. ($Gamma$ and $N$ represent the gamma and Gaussian distributions.)

@@

$$
(7)~~
\begin{align}
generalized~additive~model~(GAM): \\
f_j(t) = X_j\beta_j + Z_jb_j \\
\\
priors~for~GAM~coefficients: \\
\beta_j \sim T(0, 2.5, 3)~~[unpenalized] \\
b_j \sim N(0, \sigma_b)~~[penalized] \\
\sigma_b \sim N(0, 0.5) (seasonal~smoother) \\
\sigma_b \sim T(0, 2.5, 3) (global,~group-level~smoothers) \\ 
\end{align}
$$

where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function included in the global multi-year trend). The priors on $\sigma_b$ help to prevent overfitting by penalizing complexity in the fitted smooth terms [@mgcv]. And while the chosen priors did help with model convergence, they did not exert a strong influence on the posterior (Figure S2); the likelihood---representing `r nrow(model_in)` observations in each model---was overwhelmingly influential. A summary of the fitted models, including convergence statistics and effective sample sizes, is provided in Figure S3 and S4.

@@

## Modeling autocorrelation

@@

Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant. This general pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation. The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to nonseasonal changes in water quality over the study (see *Trends in particulate lead release*).

@@

The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the large difference in lead release between the 0 and 1 mg P L^-1^ dose. But the effect of orthophosphate dose was complex: lead release at 1 mg P L^-1^ was lower than at 0.5 mg P L^-1^ but higher than at 0.75 mg P L^-1^. We attribute the superior performance of the 0.75 mg P L^-1^ dose to an initial high dose of 2 mg P L^-1^. On the other hand, the initial 2 mg P L^-1^ dose caused excess lead release, highlighting an important trade-off that should be considered in applying a passivation-maintenance (high-low) dosing strategy. And since high orthophosphate doses may generate small lead-orthophosphate colloids/nanoparticles that are difficult to remove from drinking water [@lytle_lead_2020], the 1 mg P L^-1^ dose may be a better option in this case. The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring. This seasonal pattern may have multiple drivers: iron and water temperature, for instance, were also seasonal.
